\documentclass{article}
\input{preamble.tex}
\usepackage{pgfgantt} % for gantt diagram
\usepackage{rotating} % for rotating gantt diagram

\title{Machine Learning Project Proposal\\\textbf{Evolutionnary Computing on\\Turing machines}}
\author{Jean-Florent Raymond\\\href{mailto:Jean-Florent.Raymond.8795@student.uu.se}{\texttt{Jean-Florent.Raymond.8795@student.uu.se}}}
\date{\today}

\begin{document}
\maketitle

\section{The Busy Beaver}

\paragraph{Introduction} Given an number of states and a number of symbols, a \emph{busy beaver} is a deterministic Turing machine that, when started on a blank tape, performs ``the most operations'' before halting among all Turing machines with the same numbers of states and symbols. Performing ``the most operations'' can be definded in many ways, for instance it can mean:
\begin{itemize}
\item doing the most steps;
\item stopping with the most non-blank symbols written on the tape.
\end{itemize}

All the Turing machines we will consider here are deterministic. Let $\mathcal{T}_{m,n}$ be the set of deterministic Turing machines with an alphabet of $m$ symbols who uses $n$ states and $\mathcal{T} = \cup_{n,m \in \N} \mathcal{T}_{n,m}$.


\paragraph{Busy Beaver and Max-step Functions}
Let $\sigma : \mathcal{T} \to \N \cup \{-\infty\}$ be the function who, given a Turing machine $M$, returns:
\begin{itemize}
\item either the number of non-blank symbols that $M$, started on a blank tape, left on the tape after execution if $M$ halts;
\item or minus infinity else.
\end{itemize}

Let the \emph{busy beaver function} $\Sigma$ be defined as follows:
\[
\Sigma : \left |
\begin{array}{l}
\N \times \N \to \N\\
(m,n) \mapsto \max_{M \in \mathcal{T}_{m,n}} \sigma(M)
\end{array}
\right .
\]

Similarly, let $s : \mathcal{T} \to \N \cup \{-\infty\}$ be the function who, given a Turing machine $M$, returns
\begin{itemize}
\item either the number of steps that $M$ performs when started on a blank tape, if $M$ halts ;
\item of infinity else.
\end{itemize}

And let \emph{max-step function} $S$ be defined by
\[
S : \left |
\begin{array}{l}
  \N \times \N \to \N\\
(m,n) \mapsto \max_{M \in \mathcal{T}_{m,n}} s(M)
\end{array}
\right .
\]

Note that these four function are not computable because otherwise it would means that we can decide the halting problem \cite{rado}. Moreover there exists some relations between $S$ and $\Sigma$; for example is immediate that $S(m,n) \geq \Sigma(m,n)$, because writing a non blank on the tape require the execution of step, this becomes useful

%Maybe I would add the fact that there is no direct connection between S and \Sigma. I mean, is not possible to find the S-busy-beaver if you found for example the \Sigma-busy beaver.

\section{Finding Machines That Reaches Bounds}

The goal of this project is to find, at least for small values of $m$ and $n$, Turing machines that reaches $\Sigma(m,n)$ or $S(m,n)$, \ie finding a machine $M \in \mathcal{T}_{m,n}$ such that $\sigma(M) = \Sigma(m,n)$ or $s(M) = S(m,n)$.

Such machines have already been found \cite{rado} but only for small values of $m$ and $n$ because $S$ and $\Sigma$ functions grows very fast \cite{rado}, thing that makes tests difficult.


\paragraph{How?} This will be done using evolutionary computing techniques on a population of Turing machines coming from the same class $\mathcal{T}_{m,n}$, for fixed $m$ and $n$. The operations of crossover and mutation are made from the transition tables (the alphabet and the number of states will remain unchanged).

\paragraph{Fitness Function}
The main problem is to define a good fitness function. The value to maximize here is the number of steps before halting (for the max-step case), in order to reach the bound.

Suppose we have a machine $M$ which halts after $p$ steps and a machine $M'$ which continues to run after $p$ steps.
We cannot say that $M'$ is better w.r.t.\ max-step, because we don't know if $M'$ will stop and we have no general way to know it (since the halting problem is undecidable \cite{turing}).
However, for fixed $p \in \N$, we are able to say if a given machine will stop in less than $p$ steps. To know it, we simply run the machine during $p$ steps and watch if it is on a halting state at this time.

The approach we want to try here is depends on what bound we want to reach:
\begin{itemize}
\item for small values of $n,m \in \N$, values of $\Sigma$ and $S$ are known \cite{rado} so a suitable fitness function is a function who, given a Turing machine $M$ that we already run on $p$ steps without leading it in the halting state, returns $- \infty$ if $p > S(m,n)$ and $\frac{1}{S(m,n) - p}$ else. With this function, Turing machines which overtake the bound are highly penalized (because we know they will never halt) and a machine who run longer (but less than $S(m,n)$) without halting than an other one get a better grade.

\item for values of $n, m \in \N$ for which $\Sigma$ and $S$ are not known, we can, for growing values of $p \in \N$, find machines who halts after $p$ steps with a fitness function of the same kind as before. Then the more $p$ will be high, the more we will approximate the bound if we still find machines for this $p$.
\end{itemize}

%I agree on this distincion but.. Are we sure we can deal with the case n > 5?
%I see from the wikipedia page that for n = 6, m = 2, \Sigma already is greater or equal to 10^18267, that means we should read from the tape order of 10^18267 ones (or waiting a similarly huge number of steps), this for each Turing machine we are evaluating!
%Is my reasoning correct?

\section{Work  Packages}

The project will be split in several work packages:
\begin{enumerate}
\item Virtual Machine 
  \begin{itemize}
  \item defining a convenient way to represent Turing machines;
  \item writing a basic virtual machine to run Turing machines;
  \item adding the ability to load and save machine states in a file and to log activity;
  \item adding abilities of step counting and of running until the machine halts (if it halts).
  \end{itemize}
\item Evolutionary Computing Tool
  \begin{itemize}
  \item defining operations on Turing machines (crossover, mutations);
  \item defining a fitness function;
  \item writing a program who generates a random initial population and who follows it until it leads to something interesting.
  \end{itemize}
\item Testing
  \begin{itemize}
  \item testing all the parts of the program we write the ensure that we will not do hours of useless computation during the experiments;
  \item launching the experiments.
  \end{itemize}
\item Analyse
  \begin{itemize}
  \item analysing the results: Do we found what we were looking for? Do we discover other things?
  \item analysing the approach: Was evolutionary computing a good idea for this problem, and why?
  \end{itemize}
\item Writing documentation.
\end{enumerate}

%I agree on everything. I would try also a PSO approach if we can manage to work with the discrete search space (the binary PSO is an example).

The schedule for this project can be found on figure \ref{fig:sc}, page \pageref{fig:sc}.

\begin{sidewaysfigure}[p]
  \centering
  
  \begin{tikzpicture}[scale = 0.8]
    \begin{ganttchart}[vgrid]{37}
      \gantttitle{April}{8}\gantttitle{May}{29} \\
      \gantttitlelist{17,...,21}{7}\gantttitlelist{22}{2}\\
      \gantttitlelist[title label font={\tiny}]{23,...,30}{1}
      \gantttitlelist[title label font={\tiny}]{1,...,29}{1}\\

      \ganttgroup{Virtual Machine}{1}{19} \\
      \ganttbar{Defining the encoding}{1}{4} \\
      \ganttbar{Basic VM}{5}{13} \\
      \ganttbar{Adding functionalities}{14}{19} \\

      \ganttgroup{Evolutionary Computing Tool}{1}{19} \\
      \ganttbar{Defining the operations, etc.}{1}{3} \\
      \ganttbar{Implementing the tool}{4}{19} \\

      \ganttgroup{Testing}{1}{37} \\
      \ganttbar{Testing our code}{1}{37} \\
      \ganttbar{Launching the experiments}{20}{37} \\

      \ganttgroup{Writing Documentation}{1}{37}

    \end{ganttchart}  
  \end{tikzpicture}

  \caption{Schedule}
  \label{fig:sc}
\end{sidewaysfigure}




\section{Related Work}
The work of Shen Lin and Tibor Rado on busy bever was based on reducing the space search with normalization techniques and then on examining patterns on the tape \cite{rado}.

\bibliographystyle{acm}
\bibliography{proposal_turing}

\end{document}